Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-07-09 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added P3 journal article outline

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Understanding how search operators interact with solution representation is
a critical step to improving search.
In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
in general, the complex genotype to phenotype map makes achieving this understanding a
challenge.  By examining aspects such as tuned parameter values, the search quality of CGP variants at different
problem difficulties, node behavior, and offspring replacement properties we
seek to better understand the characteristics of CGP search.  Our focus is twofold:
creating methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
and \emph{Single}) and creating methods to overcome CGP's search limitations imposed by
genome ordering (\emph{Reorder} and \emph{DAG}).

%TODO mention this
Our results on boolean problems show CGP evolves genomes that
are highly inactive, very redundant, and full of seemingly useless constants.
On some tested problems we found less than 1\% of the genome was actually required to
encode the evolved solution.
Furthermore, traditional CGP ordering results in large portions of the genome
that are never used by any ancestor of the evolved solution.
\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
More generally, our
results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
are most likely to solve hard problems using the least number of evaluations and
the least amount of time while better avoiding degenerate
behavior.

\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Cartesian genetic programming (CGP), Analysis
\end{IEEEkeywords}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
and implement, discerning the reason that system
leads to effective search can be challenging.
This is exceptionally
true for Genetic Programming (GP), which often employs both nontrivial
genotype to phenotype maps and operators that act on genotypes without regard to
phenotypic impact.  Yet understanding the root causes of evolutionary success
and failure is critical to improving existing techniques, designing new techniques,
and understanding how and where to apply each optimization system.

There have been a number of previous studies into various aspects of GP evolution.
For instance~\cite{daida3:2003:treebias} showed how a tree's shape effects its evolvability
regardless of phenotype,
resulting in the use of new grammar based operations~\cite{xuan:2006:grammar}.
Analysis of the root cause of bloat has helped to inform methods for controlling
bloat~\cite{luke:2006:bloat}.

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering, both
%TODO final change
generally and how they specifically apply to the boolean problem domain.
%Done-NEW
While these features of CGP have trivial definitions, we will argue, using
experimentation and analysis of alternatives, that they have a great deal
of impact on search success.
%/NEW

\begin{comment}
In Section~\ref{sec:cgp} we provide
an introduction to CGP, with sufficient background to put later sections into
context.  Section~\ref{sec:duplicate} describes previous work on preventing mutation
from creating identifiably wasted evaluations.
Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
along with methods to both examine and bypass the limitations imposed by that ordering.
Each technique is then rigorously tuned and qualitatively compared on a collection
of benchmarks in Section~\ref{sec:quality}.  Section~\ref{sec:analysis} then
provides a deep look into the details of CGP's search, and how each method impacts
search, with final remarks given in Section~\ref{sec:conclusion}.
\end{comment}

\section{Cartesian Genetic Programming}
\label{sec:cgp}
Cartesian Genetic Programming (CGP) was originally proposed as a method for general
genetic programming in~\cite{miller:2000:CGPorigin}.  While its name comes from
its original application evolving circuits on a two dimensional grid, modern CGP can
represent any directed acyclic graph (DAG), and has been utilized in applications such as binary
circuits~\cite{walker:2008:cgpmodules},
robot controllers~\cite{harding:2005:robots},
neural networks~\cite{khan:2010:cgpann},
image classifiers~\cite{harding:2012:mtcgp},
and regression~\cite{harding:2009:smcgp}.

CGP represents DAGs using a linear genome of integer values.  Each node in the
DAG is encoded as a tuple of genes, with one gene specifying the function
the node applies to its inputs, and the remaining genes expressing where the
node takes input from.  Nodes can take input from either a problem input or
any node preceding them in the linear genome.  Restricting connections in this way
prevents the creation of cycles, while still allowing CGP to reuse values.  This is
in contrast to tree based GP, which must duplicate functionality everywhere the same value is needed.
To complete the representation, a set
of extra genes are included at the end of the genome to specify which nodes or
input locations to use as function outputs.

As both output locations and information flow in the DAG are evolvable, often
only a tiny fraction of the genome participates in creating the
output values.  These nodes are referred to as ``active,'' with the nodes not
being used to create output values referred to as ``inactive.''  Inactive nodes
%TODO Changed here
allow for genetic drift, as individuals can be mutated without effecting
their fitness.  These mutations can then be incorporated, as future
mutations can change the DAG structure causing previously inactive nodes to become
active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.
%Done-NEW
In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
%/NEW
each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.
%Done-NEW
This version of mutation was chosen to allow precise prediction of changes
in gene value necessarily in later sections.
%/NEW
Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

\section{Duplicate Evaluation Avoidance}
\label{sec:duplicate}
CGP's mutation operator is traditionally applied to both active and inactive genes uniformly.
This has the
potential to create offspring who's only mutated genes are in inactive nodes.
These offspring are actively identical (contain identical active genes) to their
parents, and therefore by definition have identical fitness to their parents.
As such, re-evaluating these individuals is computationally wasteful as their fitness
is known.
Detecting duplicated individuals can be done without significant overhead since
finding the set of active nodes is already done prior to
evaluation~\cite{vasicek:2012:efficient}.
Our previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
\emph{Single} as methods for avoiding or preventing these duplicate evaluations.

\begin{comment}
The number of wasted evaluations is
directly related to the mutation probability and the number of active genes.  With
sufficiently high probabilities and enough active genes, the chance of
creating an offspring which is actively identical to its parent is very low.
Yet such a configuration may make it difficult for CGP to optimize, as mutation
is likely to change many active genes in each application.  As was discussed
in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
mutation probability.
\end{comment}

\subsection{Skip}
The \emph{Skip} method for avoiding duplicate evaluations involves the least
amount of modification from CGP's normal behavior.  After an offspring is produced
and its set of active nodes is determined but before it is evaluated, it is compared
with its parent.  Each gene in each active node is compared for equivalence with
the corresponding gene in the offspring's parent.  If all active genes are found to be equal
then the offspring is not evaluated and is instead given the same fitness
as its parent.
Note that since inactive genes are not compared for equivalence it is still possible
for offspring to genetically differ from their parents.

Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
it will produce identical individuals, with a potential reduction in evaluations.
If the mutation probability is high enough relative
to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
the probability of an offspring being actively identical to its parent is effectively zero.
When the mutation probability is low enough relative to the number of active genes, some number of
offspring will be actively identical to their parents, resulting in a reduction in evaluations
but no change in the evolutionary trajectory.

%Done-NEW
This technique incurs a minor increase in runtime, as each active gene in the
offspring must be compared against the parent's gene at the same locus.
This operation is at most linear in genome size.  Yet
on problems where evaluation is even somewhat expensive, the evaluations prevented
by performing this check will result in a net reduction in runtime.
%/NEW

We would expect both intuitively and from our initial
experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
to the mutation probability than normal CGP.  We would also expect \emph{Skip} to use a lower
mutation probability than normal CGP when both are optimized to reduce evaluations.
The former comes from the fact that in normal CGP any mutation probability that has a significant
chance of creating offspring actively identical to their parents wastes
evaluations.  In \emph{Skip} there are only two penalties for reducing the
mutation probability.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
probabilities may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
exceptionally low mutation probabilities may spend a prohibitive amount of time
attempting to produce an evaluable offspring.
Note that these penalties also exist in normal CGP.

\subsection{Accumulate}
Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
creation and offspring evaluation.  Instead of skipping evaluations when offspring
are determined to be actively identical to their parents, \emph{Accumulate} enters
into a cycle of repeated mutation until an individual worth evaluating is created.
To understand this technique, consider an example where the parent $P$ produces
the offspring $F_0$ using mutation. If $F_0$ is actively identical to $P$, $F_0$
creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
$F_{i+1}$ continues until $F_n$ is produced, where $F_n$ is not actively identical
to $P$.  $F_n$ is then evaluated.  If $F_n$ is no worse than $P$, it replaces
$F_0$, and evolution continues as though $P$ had produced $F_n$ directly.  If
$F_n$ is worse than $P$, $F_{n-1}$ replaces $F_0$, as $F_{n-1}$ has accumulated
the most mutations to inactive genes without reducing fitness.  Note that
the mutation used each time is identical to normal CGP's mutation, such that
$F_n$ will have one or more active genes different from $P$.  Furthermore, even
though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
before it is evaluated, with an emphasis on encouraging neutral drift.

Previous experimentation has shown that \emph{Accumulate} acts very similarly to
\emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
favors lower mutation probabilities when tuned to reduce evaluations to success. While at first these algorithms may appear
quite different, further consideration shows how they are similar.  In \emph{Skip}
any actively identical offspring that is produced is likely to be selected as we expect mutations to
active genes to more often reduce fitness than improve it.  In the next generation
this actively identical offspring is then mutated again, with its lineage likely
continuing until it finally does mutate one or more active genes.  At this point, if the
mutant is better it replaces the parent, otherwise the parent, which has been accumulating
mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
except \emph{Skip} acts over multiple generations instead of compressed into a single generation.

%Done-NEW
Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
order to prevent wasting evaluations.  Generating each $F_i$ requires at most
$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
While \emph{Accumulate} may produce many $F_i$ in the process of creating
a single offspring, conceptually \emph{Skip} using the same mutation rate
will create a similar number of total individuals when performing the same
number of evaluations.  As such, we suggest \emph{Accumulate}
will take no more than a constant factor more runtime to produce the same number
of evaluable individuals as \emph{Skip}.  As before, this increase is likely
eclipsed when solving problems with sufficiently time consuming evaluation functions.
%/NEW

\subsection{Single}
While \emph{Skip} and \emph{Accumulate} save evaluations by not evaluating offspring
which are actively identical to their parents, \emph{Single} changes
mutation to ensure that only evaluable offspring are created.  Instead
of mutating each gene at a set probability, \emph{Single} mutates genes at
random until exactly one active gene is mutated.

This modification gives \emph{Single} three properties distinct from the other
forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
an active gene which differs from their parent.  This limits drift as the
mutated active gene must either be to an intron or represent another way to
code a solution of equal quality.  Second, as a benefit of forced changes,
\emph{Single} avoids the overhead of repeatedly generating individuals without
creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
parameter, effectively setting the mutation probability to $\frac{1}{a}$ for active
genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.

In encodings without inactive genes, limiting mutation to changing exactly one gene
could prevent an algorithm from escaping some types of local optima.  Yet because
CGP allows for inactive genes, \emph{Single} is still able to escape most local
optima using sufficient drift of inactive genes.  When a high percentage of
the genome becomes active or when \emph{Single} is otherwise limited in its ability
to drift due to lack of introns, it does have an increased potential to become stuck.
As will be discussed in Section~\ref{sec:ordering} and was shown in~\cite{goldman:2013:ordering}
this problem of highly active genomes is very unlikely.

%Done-NEW
Conceptually this method could be extended to mutate some number of active genes
before terminating.  Doing so would reintroduce a parameter to govern how many
active genes to mutate.  Our choice of a single mutation allows for the minimal
amount of change to ensure the individual should be evaluated, while mutations
made to inactive genes still give CGP sufficient power in each mutation to escape
local optimums.
%/NEW

\section{Genome Ordering}
\label{sec:ordering}
As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
to prevent cycles, by ensuring nodes only receive input from sources that
precede them in the genome.  This restriction does not limit CGP's ability
to represent DAGs, as all DAGs can be serialized to fit this requirement.
However, this representation is likely having an impact on CGP's ability to evolve specific
DAGs~\cite{goldman:2013:ordering}.

Primarily, enforcing node ordering adds artificial limitations to CGP's ability to
connect nodes.  Nodes which can be connected without creating a cycle may
still be prevented from forming that connection because of random genome ordering.
Similarly, it is impossible for adjacent nodes
to ever be connected through an intermediate node.  As a result, useful structures
in the genome may need to be evolved repeatedly in order to find the best
location for that structure in the genome.

Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
immensely inactive genomes.  As was rigorously defined in our work~\cite{goldman:2013:ordering},
the number of active genes in a genome is expected to scale logarithmically
with genome size, independent of problem application.  Furthermore, the less
active a genome, the lower the likelihood of structure ordering preventing structure
connection.

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $A$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

\begin{comment}
\begin{figure}
\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
         at a given index will be active.  $N=2000, A=2, I=6, O=6$.
         Predicts approximately 160 total active nodes, with 80\% of the first
         100 nodes active and 4\% of the remaining 1900 nodes active.}
\label{fig:prob}
\end{figure}

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.
\end{comment}
%NEW
Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob} (chosen to match
experimental values used in Section~\ref{sec:analysis})
predicts only 160 active nodes, or 8\% of the total genome.  Positionally, 80\% of the first 100 nodes are expected to be
active while just 4\% of the remaining 1900 nodes are expected to be active.
%/NEW

To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
%Done-NEW
For clarity, we shall refer to CGP's historical ordering method of enforcing
only forward connections as \emph{Normal}.
%/NEW

%DONE-NEW
It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
levels back parameter $L$ will have
a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
of the genome.  Yet this does not remove the issue of artificially limiting connections,
and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
a method which reassembled the genome using semantic cues, which circumvents the
forward only limitation in \emph{Normal}.
%/NEW
\subsection{Reorder}
\label{sec:reorder}
The \emph{Reorder} method shuffles an existing genome's node ordering without
changing node behavior.  This is possible because for a given node, there can
be a large number of other nodes which have no required ordering,
as they neither take input (directly or indirectly) from the node nor provide their
output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
locations in the genome at random once all of the nodes they take input from have been assigned
locations earlier in the genome.

The first step in performing \emph{Reorder} on a genome is to create
data structures to store direct connection relationships.  These structures are used
to determine in constant time all connections incident on a node, and 
can be built in order $O(AN)$ time, where $A$ is the arity of the nodes
and $N$ is the number of nodes.

With this preprocessing complete, the \Call{RandomSerialization}{} algorithm given in Figure~\ref{fig:reorder}
is used to assign each node a new location in the modified genome.
It starts by constructing the \emph{addable} set, which contains all nodes who's direct
dependencies have already been added.  Initially this contains only the input locations.
Once a node is assigned a location, all of the nodes that depend on its output
have that dependency marked as satisfied.  Once all of a node's dependencies
are satisfied, it can be put into \emph{addable}.  In this way, nodes are
%TODO Change here
randomly removed from \emph{addable}, assigned the next location
in the genome, with new nodes added to \emph{addable} as they becomes semantically
viable.  Iteration ends when all nodes have been assigned new locations.
\Call{RandomSerialization}{} requires $O(AN)$, as there can be only $N$ iterations,
with each iteration updating a distinct subset of the $AN$ dependencies.

\begin{figure}
  \begin{algorithmic}
    \Procedure{RandomSerialization}{}
    \State $addable \leftarrow input\_locations$
    \State $index \leftarrow |input\_locations|$
    \State $new\_loc \leftarrow \emptyset$
    \While{$|addable| > 0$}
      \State $working \leftarrow randpop(addable)$
      \If{$working \notin input\_locations$}
        \State $new\_loc[working] \leftarrow index$
        \State $index \leftarrow index + 1$
      \EndIf
      \ForAll{$link \in feeds\_to(working)$}
        \State Satisfy $link$'s dependence on $working$
        \If{$|unsatisfied(link)| = 0$}
          \State $addable \leftarrow addable \cup \{link\}$
        \EndIf
      \EndFor
    \EndWhile
    \State\Return $new\_loc$
    \EndProcedure
  \end{algorithmic}
  \caption{Algorithm which converts any DAG into a random serial ordering such
           that all nodes take input from nodes that precede them in the genome.
           Used by \emph{Reorder} to determine new node locations and \emph{DAG} to
           determine node evaluation ordering.}
  \label{fig:reorder}
\end{figure}

The final step in \emph{Reorder} is to use the list of new locations to convert
the existing genome into the reordered genome.  This can be done
by converting connection genes and output location genes using
the $new\_loc$ map returned from \Call{RandomSerialization}{}.
This also requires $O(AN)$ time, as all connection genes in all nodes must
be converted.  As this is the final step unique to \emph{Reorder}, and
no previous steps are of higher complexity, it is also the complexity of
the \emph{Reorder} algorithm as a whole.  Furthermore, it is of the same complexity class
as copying the entire genome, meaning it does not change CGP's overall complexity.

\emph{Reorder} is used once each generation to shuffle the nodes of the parent's
genome.  As the shuffling does not semantically change the parent, it does
not require re-evaluation.  Furthermore, as shuffling does not change the requirement
that nodes only depend on those preceding them in the genome, shuffling does not
require any changes to any other CGP methods, such as evaluation and mutation.
It is important to note, however, that the potential mutations that can be applied to the parent
have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
have no dependent relationship.  As $X$ does not take input from $Y$ and
$Y$ does not take input from $X$, either can be placed before the other
in the genome.  Without shuffling, the initial ordering of nodes is fixed.
No operation in \emph{Normal} is capable of making $X$
take input from $Y$.  \emph{Reorder}'s shuffling can allow
subsequent mutations to make the connection.

\emph{Reorder} has the potential to reduce node reinvention.  In \emph{Normal}, if
$X$ preceded $Y$, mutation would have to recreate either $X$ or $Y$ in new locations
in order to make the required connection.  \emph{Reorder} provides the opportunity
to examine different orderings of the genome without such costs.

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.
%Done-NEW
Therefore creating a specific offspring that improves upon the parent is
less likely, in that the number of alternatives is increased.  Yet many
of these additional alternatives may also be improvements upon the parent.  This may
increase the likelihood of producing an offspring that improves
upon the parent, allowing \emph{Reorder} to overcome this drawback.
%/NEW

\subsection{DAG}
\label{sec:dag}
\emph{Normal} and \emph{Reorder} use node location in a linear genome to create
an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
removes this restriction and allows forward and backward connections, relying on other, more complex, methodologies to
prevent cycles.  To accomplish this goal, modifications must be made to how CGP
performs mutation and evaluation.

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
%Done-NEW No change was made to mutating node as that is in fact what we wanted
%/NEW
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

\begin{figure}
  \begin{algorithmic}
  \Procedure{IsDependent}{$working$, $known$}
    \If{$working \in known$}
      \State\Return{$known[working]$}
    \EndIf
    \ForAll{$link \in reads\_from(working)$}
      \If{\Call{IsDependent}{$link$, $known$}}
        \State $known[working] \leftarrow True$
        \State\Return{$True$}
      \EndIf
    \EndFor
    \State $known[working] \leftarrow False$
    \State\Return{$False$}
  \EndProcedure
  \end{algorithmic}
  \caption{Algorithm used by \emph{DAG} to determine valid connection gene values.}
  \label{fig:dag}
\end{figure}

\Call{IsDependent}{} is a non-repeated recursive depth
first search of the individual's DAG.  Search terminates as soon as a dependent
node is found, and information from previous searches is memoized to prevent
repeated search.  Each recursion level results in a node being added to $known$.
As such initial calls will likely result in lots of recursion, but subsequent
calls will find answers more quickly.  In the worst case, this algorithm may need
to examine all nodes in the genome to determine if $working$ is dependent on the
mutating node, giving \Call{IsDependent}{} a complexity of $O(AN)$.  In the worst case, this algorithm
may be called on all possible nodes in the genome to check their dependence.
Yet cumulatively those calls can only take $O(AN)$ time due to the non-repeating
nature of the search.  As a result, the expected worst case run time for \emph{DAG}
mutation is $O(mA^2N^2)$, where $m$ is the mutation probability.  We expect $mAN$
connection genes to be mutated, each requiring a worst case of $N$ dependency checks
each taking $A$ time.
This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
mutation, as individual mutations can be performed in constant time.

The method for evaluating \emph{DAG} individuals is very similar to the method for
efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  As before,
preprocessing is done to determine which nodes are active, and only those nodes are
executed during evaluation.  Also as before, the process begins from the output locations,
recursively following connection genes and marking nodes as active until the input locations
are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
nodes that are active, but also the order in which those nodes should be executed.

\Call{RandomSerialization}{} can be reused to determine the order
in which nodes should be executed during evaluation.  Instead of using $new\_loc$
to reorder the genome, we can use this map to specify the order in which nodes should be
executed.  For instance, if a node was given the new location of $X$, we know
that once all nodes at locations preceding $X$ have been evaluated, the node at $X$
can be evaluated.  Stripping the inactive nodes from $new\_loc$ and inverting the map
results in an efficient, valid ordering in which to execute the nodes.

The process of determining which nodes are active requires $O(AN)$ time, regardless
of node ordering.  As was discussed in Section~\ref{sec:reorder},
getting the $new\_loc$ map requires $O(AN)$ time.  Finally, converting the map
into an ordering of the active nodes takes $O(N)$ time.  Each of these steps are
sequential, meaning \emph{DAG} does not increase the runtime complexity of evaluation
preprocessing.  Furthermore, once the preprocessing step is done, \emph{DAG} takes
identical time to evaluate as the other methods, regardless of the number of
input combinations.

With these modifications, there is no longer any meaning to the location of a node
in the genome.  In \emph{Normal} the number of valid
values for a connection gene is related to the number of nodes
preceding that gene, and independent of the number of nodes following that gene.
In \emph{DAG} a connection gene's valid values only depends on the current solution
represented by the individual, independent of the genes location in the genome.
As such, there are fewer artificial limitations on how \emph{DAG} can modify
solutions.

While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
Though both can create any connection that does not form a cycle,
\emph{Reorder} maintains a bias toward certain kinds of connections.
Consider again two nodes $X$ and $Y$
such that neither is dependent on the other.  If $X$ is transitively dependent on very
few nodes, and $Y$ is transitively dependent on many nodes, \emph{Reorder} is more likely
to shuffle $X$ to precede $Y$.  This is true because at any point during the construction
of the new ordering, the probability
that all of $X$'s dependencies have been added is
higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
and therefore we expect it to be added sooner.  The greater the discrepancy in the dependency
set sizes the less likely the order between the nodes will be reversed.
Conversely, \emph{DAG} has no such bias.  It is always possible to mutate $X$ to connect to $Y$.

The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
For example, assume $X$ has more nodes transitively dependent on it than $Y$.
In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
have a higher chance of connecting to $Y$ than $Y$ to $X$.  This is because $X$ can connect to fewer
nodes, so each mutation has a higher chance of choosing $Y$.

\section{Qualitative Comparisons}
\label{sec:quality}

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
TODO Abstract
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
TODO
\end{IEEEkeywords}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{TODO}

~\cite{Wolpert:1997:nfl}

~\cite{goldberg:1991:gasize} % Search efficiency

~\cite{grefenstette:1986:optimalga} % Expensive to tune

~\cite{harik:1999:parameterlessga}

~\cite{posik:2011:parameterless} % Model based parameter-less, only noiseless non-overlapping

~\cite{goldman:2014:p3}

\section{Optimizers}

\subsection{P3}
~\cite{goldman:2014:p3}

~\cite{hornby:2006:alps}

~\cite{gronau:2007:upgma} % Efficient clustering

\subsection{Hill Climber}
~\cite{goldman:2014:p3}

\subsection{LTGA}
~\cite{thierens:2010:ltga}

~\cite{thierens:2013:ltgahiff}

~\cite{bosman:2011:lsbbo} % LTGA with hill climbing

\subsection{hBOA}

\subsection{Parameter-less hBOA}
~\cite{pelikan:2004:parameterlesshboa}

\subsection{$1+(\lambda, \lambda)$}
~\cite{doerr:2013:lambdalambda}

~\cite{goldman:2014:p3}

\section{Problem Descriptions}
\subsection{Single Instance Problems}
~\cite{goldberg:1991:gasize} % Deceptive Trap

~\cite{goldman:2012:ltga} % Deceptive Step Trap

~\cite{thierens:2013:ltgahiff} % HIFF

~\cite{goldman:2014:p3} % Rastrigin

\subsection{Randomly Generated Problem Classes}

~\cite{wright:2000:solvingnk}

~\cite{saul:1994:spinglass}
\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}

~\cite{goldman:2014:p3} % MAXSAT

\section{Comparison Algorithm Parameter Tuning}
~\cite{goldman:2012:ltga} % Bisection with no parameters

~\cite{goldman:2014:p3}

~\cite{jovanovic:1997:ruleofthree}

\section{Global Optimum}
TODO Compare all 6 optimization techniques on time to reach global optimum

TODO Table listing empirical $O(N)$ times for each technique on each problem. Include
statistical comparison of log-log regression lines.

TODO Consider a graphic showing 5\% and 95\% Quantiles to illustrate P3's scaling on problems

\section{Stop Anytime}
TODO Compare all 6 optimization techniques on fitnesses reached at different times
during the run on problem sizes near 100 or 150

TODO Compare all 4 optimization techniques on fitnesses reached at different times
during the run on problem sizes near 500 or 750

\section{Complexity Per Evaluation}
TODO Include graphic of rebuilds per evaluation (decreasing with length)

TODO Include graphic of donations per evaluation (sub-linear growth)

TODO Include graphic comparing seconds per evaluation up to small problems (hboa super-linear)

TODO Include graphic comparing seconds per evaluation for large problems (perhaps just 1 graph of P3 across all problems)

\section{Inner Workings}
TODO Include graphic of proportion of evaluations spend on hill climber (constant with length)

TODO Include graphic of crossover success (constant with length)

TODO Include graphic of Ising Spin Glass, NK, and DeceptiveStepTrap's pyramid shape and crossover success rates at each level

~\cite{lobo:2011:dynamicpop}

\section{Conclusions and Future Work}
% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed
